Designing Types for R, Empirically

### 2020-05-16 paper/main.tex
Typos

Our community builds, improves, and reasons about programming languages.  To
make design decisions that benefit users, we need to understand our target
language as well as the real-world needs it answers. Often, we can appeal to
our intuition, as many languages are intended for general purpose
programming tasks. Unfortunately, intuition may fail when looking at
domain-specific languages as these languages designed for a specific group
of users to solve very specific needs. This is the case of the data science
language R.

@@

Many of the design decisions that gave us R were intended to foster an
interactive, exploratory, programming style. These include, to name a few,
the lack of type annotations, the ability to use syntactic shortcut, and the
automatic conversion between data types.  While these choices have led to a
language with a low barrier to entry---many data science educational
programs do not teach R itself but simply introduce some of its key
libraries---they have also created a language where errors can go
undetected.

Retrofitting a type system to the R programming language would increase our
assurance in the result of data analysis. But, we are faced with two
challenges. First, it is unclear what would be the \emph{right} type system
for a language as baroque as R. For example, one of the most popular data
type, the \code{data.frame}, is manipulated through reflective operations --
a data frame is a table whose columns can be added or removed on the fly.
Second, but just as crucially, designing a type system that will be adopted
would require overcoming some prejudices and educating large numbers of
users.

@@

One way to validate the inferred type annotations is to check that different
programs using one of the functions for which a type signature has been
inferred respect the signature.  \contractr is an R package that allows to
decorate function with assertions. We use it to insert type checks of the
arguments and return values. As usual with R, this is not straightforward.

While \contractr's primary logic has been implemented in C++ to reduce its
runtier overhead, we have not observed a single segfault during our use of
the package.  It has been tested with GNU R-3.5.0 and hardened with a
battery of 400 test cases. It works by modifying function definition to
insert a call to a type-checking function.

@@

We have downloaded and installed all available CRAN packages. Out of the
\AllCranRnd packages, we managed to install \AllLoadableRnd. The main reason for
this is that \rdt is based on R 3.5.0 and some of the packages are no longer
compatible with this version. Some packages also require extra native
dependencies which were not present on our servers. We defined two criteria for
including a package into the corpus:
\begin{inparaenum}[(1)]
  \item have a runnable code that covers a significant part of the package
  source code from which type signatures could be inferred, and
  \item have some reverse dependencies that will allows us to evaluate the
  inferred types using the runnable code from these dependencies.
\end{inparaenum}
The concrete thresholds used were: at least \ThresholdCodeCoverage of expression
coverage and at minimum \ThresholdRevdeps reverse dependencies. The code
coverage was computed for each package using
\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R tool.
The reverse package dependencies were extracted from the package metadata using
builtin function.

@@

How robust are the inferred types?  To measure this, we conducted another
large-scale experiment: for each package in the corpus, we ran used the
inferred type signatures as contracts and ran all of the CRAN reverse
dependencies for that package. In total we ran \NUMPKGSEVAL packages and
recorded \NUMASSERTIONS total assertions.  Overall, we found that only
\PERCFAILEDASSERTIONSWUNDEF of contract failed.  The limit on number of
arguments (we record only 20) accounted for \PERCASSERTIONSUNDEF failed
assertion. We found that \PERCSUCCARG of parameter types and \PERCSUCCFUNS
of functions types never failed.  The number of immaculate function types
increases to \PERCSUCCFUNNOSTHREE if we discount S3 object method dispatch.
Overall, these numbers are promising, and suggest that the type signatures
are robust.

@@

\begin{table}
% BEGIN Autogenerated
\small
\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & Occurrences & \% Total & Cumulative \%\\
\midrule
\VEC\dbl & \CLASS{\texttt{matrix}} & 371,896 & 27.37 & 27.37\\
\VEC\dbl & \VEC\int & 117,212 & 8.62 & 35.99\\
\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 100,100 & 7.37 & 43.36\\
\dbl & \int | \null & 60,578 & 4.46 & 47.81\\
\VEC\dbl & \CLASS{\texttt{timeSeries}} & 58,872 & 4.33 & 52.15\\
\addlinespace
\CLASS{\texttt{matrix}} & \CLASS{\texttt{timeSeries}} & 55,005 & 4.05 & 56.19\\
\VEC\dbl & \dbl & 53,338 & 3.92 & 60.12\\
\dbl & \CLASS{\texttt{data.frame}} & 32,553 & 2.40 & 62.51\\
\VEC\dbl & \CLASS{\texttt{data.frame}} & 31,576 & 2.32 & 64.84\\
\dbl & \int & 20,586 & 1.51 & 66.35\\
\bottomrule
\end{tabular}
% END Autogenerated
\label{tbl:fail-breakdown-0}\caption{Top contract failures}\end{table}

@@

% END Autogenerated
 \label{tbl:fail-breakdown-1}
 \caption{Results from Table 4, broken down by occurrences of the expected type as a parameter type}
 \end{table}

@@

 \begin{table}
% BEGIN Autogenerated\
\small
\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
\VEC\dbl & \VEC\int & 44 & 657 & 6.70\\
\NAVEC\lgl & \null & 158 & 2756 & 5.73\\
\VEC\chr & \chr & 183 & 4102 & 4.46\\
\chr & \null & 117 & 2756 & 4.25\\
\dbl & \null & 107 & 2756 & 3.88\\
\bottomrule
\end{tabular}
% END Autogenerated
 \label{tbl:fail-breakdown-2}
 \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 90th percentile}
 \end{table}

  \begin{table}
% BEGIN Autogenerated
\small
\begin{tabular}{ll@{}rrr@{}}
\toprule
Passed & Arg Type & \#Args Fail & \#Args w Type & \% Failure\\
\midrule
\CLASS{\texttt{data.frame}} & \CLASS{\texttt{data.frame}, \texttt{tbl}, \texttt{tbl\_df}} & 63 & 284 & 22.18\\
\CLASS{\texttt{matrix}} & \CLASS{\texttt{array}} & 13 & 70 & 18.57\\
\NAVEC\lgl & \lgl | \null & 9 & 49 & 18.37\\
\VEC\chr & \LIST{\VEC\chr} & 20 & 109 & 18.35\\
\VEC\clx & \clx & 7 & 42 & 16.67\\
\bottomrule
\end{tabular}
% END Autogenerated
 \label{tbl:fail-breakdown-3}
 \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 80th percentile}
 \end{table}

>>>>

Our community builds, improves, and reasons about programming languages. To make
design decisions that benefit users, we need to understand our target language
as well as the real-world needs it answers. Often, we can appeal to our
intuition, as many languages are intended for general purpose programming tasks.
Unfortunately, intuition may fail when looking at domain-specific languages as
these languages were designed for a specific group of users to solve very
specific needs. This is the case of the data science language R.

@@

Many of the design decisions that gave us R were intended to foster an
interactive and exploratory programming style. These include, to name a few,
the lack of type annotations, the ability to use syntactic shortcut, and the
automatic conversion between data types. While these choices have led to a
language with a low barrier to entry---many data science educational
programs do not teach R itself but simply introduce some of its key
libraries---they have also created a language where errors can go
undetected.

Retrofitting a type system to the R programming language would increase our
assurance in the result of data analysis. But, we are faced with two
challenges. First, it is unclear what would be the \emph{right} type system
for a language as baroque as R. For example, one of the most popular data
type, the \code{data.frame}, is manipulated through reflective operations---a data frame is a table whose columns can be added or removed on the fly.
Second, but just as crucially, designing a type system that will be adopted
would require overcoming some prejudices and educating large numbers of
users.

@@

One way to validate the inferred type annotations is to check that different
programs using one of the functions for which a type signature has been inferred
respect the signature. For this purpose, we developed \contractr, an R package
that allows one to decorate function with assertions. We use it to insert type
checks of the arguments and return values. As usual with R, this is not
straightforward.

While \contractr's primary logic has been implemented in C++ to reduce its
runtime overhead, we have not observed a single segfault during our use of
the package.  It has been tested with GNU R-3.5.0 and hardened with a
battery of 400 test cases. It works by modifying function definition to
insert a call to a type-checking function.

@@

We have downloaded and installed all available CRAN packages. Out of the
\AllCranRnd packages, we managed to install \AllLoadableRnd. The main reason for
this is that \rdt is based on GNU-R 3.5.0 and some of the packages are no longer
compatible with this version. Some packages also require extra native
dependencies which were not present on our servers. We defined two criteria for
including a package into the corpus:
\begin{inparaenum}[(1)]
  \item have a runnable code that covers a significant part of the package
  source code from which type signatures could be inferred, and
  \item have some reverse dependencies that will allows us to evaluate the
  inferred types using the runnable code from these dependencies.
\end{inparaenum}
The concrete thresholds used were: at least \ThresholdCodeCoverage of expression
coverage and at minimum \ThresholdRevdeps reverse dependencies. The code
coverage was computed for each package using
\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R tool.
The reverse package dependencies were extracted from the package metadata using
builtin function.

@@

How robust are the inferred types? To measure this, we conducted another
large-scale experiment: for each package in the corpus, using the inferred type
signatures as contracts we ran all of the CRAN reverse dependencies for that
package. In total we ran \NUMPKGSEVAL unique packages and recorded
\NUMASSERTIONS total assertions. Overall, we found that only
\PERCFAILEDASSERTIONSWUNDEF of contract failed. The limit on number of arguments
(we record only 20) accounted for \PERCASSERTIONSUNDEF failed assertion. We
found that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of functions types
never failed. The number of immaculate function types increases to
\PERCSUCCFUNNOSTHREE if we discount S3 object method dispatch. Overall, these
numbers are promising, and suggest that the type signatures are robust.

@@

\begin{table}
% BEGIN Autogenerated
\small
\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & Occurrences & \% Total & Cumulative \%\\
\midrule
\VEC\dbl & \CLASS{\texttt{matrix}} & 371,896 & 27.37 & 27.37\\
\VEC\dbl & \VEC\int & 117,212 & 8.62 & 35.99\\
\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 100,100 & 7.37 & 43.36\\
\dbl & \int | \null & 60,578 & 4.46 & 47.81\\
\VEC\dbl & \CLASS{\texttt{timeSeries}} & 58,872 & 4.33 & 52.15\\
\addlinespace
\CLASS{\texttt{matrix}} & \CLASS{\texttt{timeSeries}} & 55,005 & 4.05 & 56.19\\
\VEC\dbl & \dbl & 53,338 & 3.92 & 60.12\\
\dbl & \CLASS{\texttt{data.frame}} & 32,553 & 2.40 & 62.51\\
\VEC\dbl & \CLASS{\texttt{data.frame}} & 31,576 & 2.32 & 64.84\\
\dbl & \int & 20,586 & 1.51 & 66.35\\
\bottomrule
\end{tabular}
% END Autogenerated
\caption{Top contract failures}
\label{tbl:fail-breakdown-0}
\end{table}

@@

% END Autogenerated
 \caption{Results from Table 4, broken down by occurrences of the expected type as a parameter type}
 \label{tbl:fail-breakdown-1}
 \end{table}

@@

 \begin{table}
% BEGIN Autogenerated\
\small
\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
\VEC\dbl & \VEC\int & 44 & 657 & 6.70\\
\NAVEC\lgl & \null & 158 & 2756 & 5.73\\
\VEC\chr & \chr & 183 & 4102 & 4.46\\
\chr & \null & 117 & 2756 & 4.25\\
\dbl & \null & 107 & 2756 & 3.88\\
\bottomrule
\end{tabular}
% END Autogenerated
 \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 90th percentile}
 \label{tbl:fail-breakdown-2}
 \end{table}

  \begin{table}
% BEGIN Autogenerated
\small
\begin{tabular}{ll@{}rrr@{}}
\toprule
Passed & Arg Type & \#Args Fail & \#Args w Type & \% Failure\\
\midrule
\CLASS{\texttt{data.frame}} & \CLASS{\texttt{data.frame}, \texttt{tbl}, \texttt{tbl\_df}} & 63 & 284 & 22.18\\
\CLASS{\texttt{matrix}} & \CLASS{\texttt{array}} & 13 & 70 & 18.57\\
\NAVEC\lgl & \lgl | \null & 9 & 49 & 18.37\\
\VEC\chr & \LIST{\VEC\chr} & 20 & 109 & 18.35\\
\VEC\clx & \clx & 7 & 42 & 16.67\\
\bottomrule
\end{tabular}
% END Autogenerated
 \caption{What is the highest failure rate among popular argument types? For argument signatures whose frequency is in the 80th percentile}
 \label{tbl:fail-breakdown-3}
 \end{table}
